Composite Bayesian inference

### 2018-11-13 papers/composite_inference/draft.tex
Modifs on composite inference paper

\graphicspath{{pics/}}

@@

\title{Composite Bayesian inference}
\date{May 3, 2016}

\author{Alexis Roche\thanks{alexis.roche@\{centraliens.net,gmail.com,epfl.ch\}}}
%\\
%CHUV / Siemens Healthcare / EPFL \\ 
%CH-1015 Lausanne, Switzerland}

@@

\begin{abstract}
This paper revisits the concept of composite likelihood from the perspective of probabilistic inference, and proposes a generalization called {\em super composite likelihood} for sharper inference in multiclass problems. It is argued that, beside providing a new interpretation and a general justification of na\"ive Bayes procedures, super composite likelihood yields a much wider class of discriminative models suitable for unsupervised and weakly supervised learning.
\end{abstract}

@@

Conventional Bayesian inference and other likelihood-based paradigms rest upon the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable. Because this is challenging when the data is inherently complex, common practice to implement feasible inference algorithms is to use deliberately misspecified data-generating models \cite{White-82,Walker-13} such as in {\em na\"ive Bayes} \cite{Ng-01} or minimum description length \cite{Grunwald-07}, or to resort to highly supervised discriminative modeling approaches \cite{Ng-01}, not to mention ad-hoc methodology.

Composite likelihood (see \cite{Varin-11} and the references therein) is a middle-way approach that extends the familiar notion of likelihood without requiring a full data-generating model. The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole. This may also be viewed as a {\em divide \& conquer} method of approximating the actual likelihood. While maximum composite likelihood does not inherit the general property of maximum likelihood to yield asymptotically minimum-variance estimators, it may offer an excellent trade-off between computational and statistical efficiency.

In this note, composite likelihood is interpreted as a probabilistic opinion pool \cite{Genest-86,Garg-04} of ``agents'' using different pieces of information, or clues, extracted from the data. Each agent acts as a local Bayesian statistician expressing an opinion in the form of a posterior distribution on the unknown parameters of interest, or hypotheses, given a specific clue. Composite likelihood can therefore be associated with a probability distribution on hypotheses, hence extending Bayesian analysis to problems where the proper likelihood function is intractable.

I further justify a generalization of composite likelihood called {\em super composite likelihood} whereby clues can be weighted differently depending on hypotheses to better deal with multiclass inference problems. This more general concept also encompasses another likelihood approximation strategy known as {\em PDF~projection} \cite{Baggenstoss-03,Minka-04,Baggenstoss-15}.

This paper is intended to serve as supporting material for a companion paper (in preparation), where the super composite likelihood framework is applied in computer vision to develop a probabilistic theory of image registration \cite{Viola-97}.

@@

Let $Y$ an observable multivariate random variable with sampling distribution $p(y|\theta)$ conditional on some unobserved parameter of interest, $\theta\in\Theta$, where $\Theta$ is a known set. Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $\theta$:
$$
L(\theta) = p(y|\theta)
.
$$

For a high-dimensional $Y$, this expression may be intractable if a plausible generative model is lacking, or involves nuisance parameters that are cumbersome to integrate out. A natural workaround known as {\em data reduction} is then to extract some lower-dimensional feature $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
$$
\ell(\theta) = p(z|\theta)
.
$$

Substituting $L(\theta)$ with $\ell(\theta)$ boils down to restricting the sample space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information. While it is valid for such an agent observing~$z$ only to consider $\ell(\theta)$ as the likelihood function of the problem, the drawback is that $\ell(\theta)$ might be poorly informative about~$\theta$ due to the information loss incurred by data reduction. To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to aggregate the likelihood functions $\ell_i(\theta) = p(z_i|\theta)$ that they elicit.

This leads to a classical problem of combining probabilistic opinions from possibly redundant agents \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. Genest {\em et al} \cite{Genest-86b} showed, in particular, that the only pooling operator that does not explicitly depend on~$\theta$ and preserves {\em external Bayesianity} is the generalized logarithmic opinion pool:
\begin{equation}
\label{eq:log_pool}
p_\star(\theta) = \frac{1}{Z} \pi(\theta) \prod_{i=1}^n \ell_i (\theta)^{w_i},
\qquad
{\rm with}\quad 
Z = \int \pi(\theta) \prod_{i=1}^n \ell_i (\theta)^{w_i} d\theta,
\end{equation} 
where $\pi(\theta)$ is some reference distribution or prior, and $w_i$ are arbitrary positive\footnote{Negative weights can be chosen only if the parameter set $\Theta$ is finite \cite{Genest-86b}.} weights that sum up to one,
$$
\sum_{i=1}^n w_i = 1
.
$$

External Bayesianity essentially means that it should not matter whether a prior on~$\theta$ is incorporated before or after pooling opinions, provided that all agents agree on the same prior. Importantly, the log-linear pool does not assume mutual feature independence, as would the same factorized form as (\ref{eq:log_pool}) with all unitary weights, $w_1=\ldots=w_n= 1$. Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature. This results in increasing their weights in the log-linear pool.}.

Strikingly, (\ref{eq:log_pool}) reduces to an analogous of Bayes rule: $p_\star(\theta)\propto \pi(\theta) L_c(\theta,\mathbf{w})$, where $\mathbf{w}=(w_1,w_2,\ldots,w_n)^\top$ denotes the vector of weights, and the quantity:
\begin{equation}
\label{eq:comp_lik}
L_c(\theta,\mathbf{w}) \equiv \prod_{i=1}^n \ell_i (\theta)^{w_i}
\end{equation} 
plays exactly the same role as a traditional likelihood function. $L_c(\theta,\mathbf{w})$ shares a convenient factorized form with the likelihood derived under mutual feature independence, sometimes called {\em na\"ive Bayes} likelihood in the literature \cite{Ng-01}. The key difference is that the single-feature likelihoods are scaled by positive weights~$w_i$ smaller than one, hence producing a flatter posterior distribution. In comparison with the proper likelihood (not assuming feature independence), the clear computational advantage is that we only need to evaluate the marginal feature distributions, rather than the joint distribution of all features.

We recognize in~(\ref{eq:comp_lik}) a general expression known as a {\em marginal composite likelihood} \cite{Varin-11}, although it is derived here under the restriction that the weights sum up to one (as already motivated in \cite{Wang-14} by a different argument using the maximum entropy principle). 
% Accounting here for Xi'An comment on xianblog.wordpress.com
% The sum of the powers is constrained to be equal to one, even though
% I do not understand why the dimensions of the projections play no
% role in this constraint. Simplicity is advanced as an argument,
% which sounds rather weak…
This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $\theta$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.

Previous attempts at Bayesian composite likelihood include tuning a constant weight so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or performing a close-in-spirit curvature adjustment \cite{Ribatet-12}. Such approaches reconcile the frequentist and Bayesian notions of uncertainty to some extent, but are not externally Bayesian since they do not warrant unit sum weights. When we refer to composite likelihood in the sequel, we assume unit sum weights.

%It is an open problem to determine weights that make it appropriate to plug composite likelihood into Bayes rule. The recommendation for unit sum weights is particularly simple compared to, {\em e.g.}, tuning a constant weight so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or performing a close-in-spirit curvature adjustment \cite{Ribatet-12}. Such approaches reconcile the frequentist and Bayesian notions of uncertainty to some extent, but are not externally Bayesian since they do not warrant unit sum weights. When we refer to composite likelihood in the sequel, we assume unit sum weights.

@@

Composite likelihood may also be understood as a means to approximate the ``true'' likelihood function\footnote{By {\em true likelihood}, we mean the likelihood corresponding to a specified yet intractable generative model. In practice, such a model is obviously not required.} or, in the language of graphical models, the {\em message} that the data sends to the latent variable~$\theta$. Several ``clues'' are sent to~$\theta$ via different data features, and then integrated assuming {\em non-coalescent} emitting sources, meaning that statistical dependences between clues are treated as unknown, although not ignored. The whole idea is depicted by the factor graph\footnote{See \cite{Bishop-06} for a didactic introduction to factor graphs.} in Figure~\ref{fig:fgraph2}, where the latent variable is connected to multiple factors involving single clues (rather than a single factor involving multiple clues), thereby enabling efficient computations.

@@

\begin{figure}[!ht]
\begin{center}
\subfigure[Generative model]{\includegraphics[width=.35\textwidth]{fgraph1.pdf}\label{fig:fgraph1}}
\subfigure[Composite likelihood model]{\includegraphics[width=.6\textwidth]{fgraph2.pdf}\label{fig:fgraph2}}
\caption{Factor graph representations of a generative model (a) and its approximation using composite likelihood (b). In (a), the factor connecting the data~$y$ and the latent variable~$\theta$ is the true likelihood, $L(y,\theta)=p(y|\theta)$. In (b), factors connecting~$y$ and the clues $z_1,z_2,\ldots$ represent feature extractions, $\alpha_i(y,z_i)=\delta[z_i-f_i(y)]$, while factors connecting the clues and~$\theta$ are scaled single-feature likelihoods, $\beta_i(z_i,\theta)=p(z_i|\theta)^{w_i}$. In both graphs, the factor shown in black is the prior $\pi(\theta)$.}
\label{fig:fgraph}
\end{center}
\end{figure}

@@

If one interprets the average divergence from agent opinions~(\ref{eq:avg_kl}) as a proxy for the divergence from ``the truth'', $p(\theta|y)$, composite likelihood can be seen as a variational approximation to the true likelihood. This corresponds to the intuitive notion that a consensus among sufficiently many experts should yield a reasonable guess. An essential difference with usual approximate inference methods \cite{Bishop-06,Minka-05} is that the likelihood function does not need to be computable owing to the use of a proxy. However, as we shall see in Section~\ref{sec:weights}, knowledge of the feature sampling distributions can be further exploited to optimize the likelihood approximation with respect to the composite likelihood weights.

\section{Super composite likelihood}
\label{sec:super}

Due to the distribution of weights between clues, a drawback of composite likelihood is that it is prone to {\em information overload} in the sense that it tends to ``flatten out'' when too many clues are included as relevant clues then get downweighted. If one decides not to merge clues for computational reasons (this would require handling their joint distribution), one could hope to mitigate information overload by assigning strong weights only to those clues that are believed to be ``most informative''.

However, when chosen for computational simplicity, clues may not only convey limited information at individual level: their informativeness may also be very much hypothesis-dependent. Consider, for instance, diagnosing a disease from a routine medical checkup. Body temperature may point to a bacterial infection by comparison with normality, but would not help detecting a non-infectious cardiovascular disease -- and conversely for, say, blood pressure.

This motivates a more general setting where clues can be weighted differently depending on hypotheses. To avoid unnecessary technicalities, we will assume from now on a finite set of hypotheses, $\Theta=\{\theta_0,\theta_1,\ldots,\theta_m\}$, where one particular hypothesis, $\theta_0$, is given the special status of reference, or ``null'' hypothesis.

We start by introducing auxiliary binary variables~$t_j$, for $j=1,\ldots,m$, defined by truncation of~$\theta$:
\begin{equation}
\label{eq:binary}
t_j = 
\left\{
\begin{array}{lll}
1 & {\rm if} & \theta = \theta_j \\
0 & {\rm if} & \theta = \theta_0
\end{array}
\right.
.
\end{equation}
One can think of each~$t_j$ as an indicator light that flashes green or red whenever~$\theta$ is in one of the particular two states $\theta_j$ or $\theta_0$, and does not respond otherwise. The collection of all $t_j$'s may be thought of as a population code \cite{Knill-04} for~$\theta$.

The key idea is as follows: instead of approximating the message sent from the data to~$\theta$ in one piece as in Section~\ref{sec:message}, we may approximate each of the simpler messages sent to the $t_j$'s. To that end, we construct a factor graph using the population code, depicted in Figure~\ref{fig:fgraph3}, which is equivalent to the graph in Figure~\ref{fig:fgraph1} for the purpose of computing the posterior distribution~$p(\theta|y)$. This graph involves multiple factors representing the ``truncated'' likelihood functions, 
$$
L_j(t_j)=p(y|t_j)=
\left\{
\begin{array}{lll}
p(y|\theta_j) & {\rm if} & t_j=1 \\
p(y|\theta_0) & {\rm if} & t_j=0
\end{array}
\right.
,
\qquad j=1,\ldots,m,
$$
as opposed to a single factor for the full likelihood $L(\theta)$. In addition, there are factors $\gamma_j$ to synthesize the different messages sent to the binary variables $t_1,\ldots,t_m$ into one sent to~$\theta$:
\begin{equation}
\label{eq:factor_gamma}
\gamma_j(t_j,\theta)
= 
\left\{
\begin{array}{lll}
\delta_{1t_{j}} & {\rm if} & \theta=\theta_j \\
\delta_{0t_{j}} & {\rm if} & \theta\not=\theta_j
\end{array}
\right.
,
\end{equation}
where $\delta$ denotes the Kronecker delta.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.6\textwidth]{fgraph3.pdf}
  \end{center}
\caption{Alternative graph using population coding for the computation of $p(\theta|y)$. See text for details.}
\label{fig:fgraph3}
\end{figure}

The unnormalized joint distribution~$p_2(y,\theta)$ represented by the graph in Figure~\ref{fig:fgraph3} reads:
\begin{eqnarray*}
p_2(y,\theta) 
& = & 
\pi(\theta) \sum_{t_1=0}^{1}\ldots \sum_{t_m=0}^{1} 
\prod_{j=1}^m L_j(t_j) \gamma_j(t_j,\theta) \\
& = & 
\pi(\theta)  
\prod_{j=1}^m \left\{ 
\mathbf{1}_{\{\theta_j\}}(\theta) p(y|\theta_j) + [1-\mathbf{1}_{\{\theta_j\}}(\theta)]p(y|\theta_0)
\right\}\\
& = & 
\pi(\theta) p(y|\theta) p(y|\theta_0)^{m-1}
,
\end{eqnarray*}
clearly yielding the same posterior $p_2(\theta|y)=p(\theta|y)$ as the graph in Figure~\ref{fig:fgraph1}, although generally not the same generative model: $p_2(y|\theta)\propto p(y|\theta)p(y|\theta_0)^{m-1}$ hence $p_2(y|\theta)\not=p(y|\theta)$ unless there are two hypotheses only ($m=1$), or the null distribution~$p(y|\theta_0)$ is uniform. While the inconsistency between generative models is irrelevant to inference, the alternative graph enables a more flexible likelihood approximation scheme, where each factor $L_j(t_j)$ can be substituted with a specific composite likelihood $L_{cj}(t_j, \mathbf{w}_j)$ using own pre-determined weights $\mathbf{w}_j=(w_{1j},w_{2j},\ldots,w_{nj})^\top$ depending on~$j$:
$$
L_{cj}(t_j, \mathbf{w}_j) 
= \prod_{i=1}^n p(z_i|t_j)^{w_{ij}}
= \prod_{i=1}^n \beta_{ij}(z_i, t_j),
$$
with:
\begin{equation}
\label{eq:factor_beta}
\beta_{ij}(z_i, t_j)
= 
\left\{
\begin{array}{lll}
p(z_i|\theta_j)^{w_{ij}} = \ell_i (\theta_j)^{w_{ij}} & {\rm if} & t_j=1 \\
p(z_i|\theta_0)^{w_{ij}} = \ell_i (\theta_0)^{w_{ij}} & {\rm if} & t_j=0
\end{array}
\right.
,
\end{equation}
and:
$$
\forall j \in \{1,\ldots,m\},
\qquad
\sum_{i=1}^n w_{ij} = 1
.
$$

%Let $\mathbf{W}$ denote an $n\times m$ matrix of pre-determined unit sum weights $w_{ij}$ such that each column, denoted $\mathbf{w}_j$, sums up to one. The possibly sparse $\mathbf{W}$ encodes prior knowledge about the relative relevance of the selected clues to the different hypotheses under investigation.

This corresponds to replacing each factor $L_j$ in Figure~\ref{fig:fgraph3} with a subgraph of same structure as the subgraph highlighted in red in Figure~\ref{fig:fgraph2}, resulting in the further modified factor graph shown in Figure~\ref{fig:fgraph4}. Intuitively, hypothesis-dependent weights make it possible to emphasize the clues that are relevant to each particular hypothesis comparison $\theta_j$ {\em vs.}~$\theta_0$, leading to potentially better approximations to the true odds $p(\theta_j|y)/p(\theta_0|y)$ than using constant weights.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.7\textwidth]{fgraph4.pdf}
  \end{center}
\caption{Super composite likelihood factor graph. See respectively (\ref{eq:factor_beta}) and (\ref{eq:factor_gamma}) for the expression of the factors $\beta_{ij}$ and  $\gamma_j$.}
\label{fig:fgraph4}
\end{figure}

We shall point out that the subgraph connecting ${\bf z}=(z_1,z_2,\ldots,z_n)$ and ${\bf t}=(t_1,t_2,\ldots,t_m)$ in Figure~\ref{fig:fgraph4} (shown in red) is equivalent to an undirected bipartite graph, a property shared with restricted Boltzmann machines (RBM) \cite{Fischer-14,Hinton-06,Larochelle-08}. This means that the variables in one layer are independent conditionally on the other layer:
\begin{equation}
\label{eq:bipartite}
p_3({\bf t}|{\bf z}) = \prod_{j=1}^m p_3(t_j|{\bf z}),
\qquad
p_3({\bf z}|{\bf t}) = \prod_{i=1}^n p_3(z_i|{\bf t}),
\end{equation}
with, in this case, 
$$
p_3(t_j|{\bf z}) \propto L_{cj}(t_j, \mathbf{w}_j),
\qquad
p_3(z_i|{\bf t}) = p_3(z_i|\theta) \propto p(z_i|\theta)^{W_i(\theta)},
$$
where $W_i(\theta)$ is defined by $W_i(\theta_j)=w_{ij}$ if $j\in\{1,m\}$ and $W_i(\theta_0)=\sum_{j=1}^m w_{ij}$.

% and may hence be considered as a kind of restricted Boltzmann machine (RBM) \cite{Fischer-14,Hinton-06,Larochelle-08}: specifically, one with real-valued (or even vector-valued) inputs and truncated binary outputs. The parameters of this RBM are the feature sampling distribution parameters as well as the weights~${\bf W}$. Assuming that the former are known, there is a very natural method to learn~${\bf W}$, as we will see in Section~\ref{sec:weights}, which differs from traditional RBM~learning methods, namely contrastive and discriminative learning.

An essential difference with RBM, however, is that the generative distribution $p_3({\bf z}|{\bf t})$ is not intended to be used for model training. Instead, our core assumption is that the marginal feature distributions are learned ``before'' constructing the graph, a step that does not rely on assuming feature independence conditionally on~${\bf t}$ or, equivalently, $\theta$. Therefore, despite satisfying the bipartite condition (\ref{eq:bipartite}), the joint distribution $p_3({\bf t},{\bf z})$ represented in Figure~\ref{fig:fgraph2} is asymmetrical in use, and serves the only purpose of defining a posterior distribution~$p_3(\theta|y)$ via~$p_3({\bf t}|{\bf z})$.

Marginalizing out~${\bf t}$, we find:
\begin{eqnarray*}
p_3(\theta|y) 
& \propto & 
\pi(\theta) \sum_{{\bf t}\in \{0,1\}^m} 
\prod_{j=1}^m p_3({\bf t}|{\bf z}) \gamma_j(t_j,\theta) \\
& = & 
\pi(\theta) \sum_{t_1=0}^{1}\ldots \sum_{t_m=0}^{1} 
\prod_{j=1}^m L_{cj}(t_j, \mathbf{w}_j) \gamma_j(t_j,\theta) \\
& = & 
\pi(\theta) 
\prod_{i=1}^n
\prod_{j=1}^m
\ell_i \left(
\mathbf{1}_{\{\theta_j\}}(\theta) \theta + [1-\mathbf{1}_{\{\theta_j\}}(\theta)]\theta_0
\right)^{w_{ij}} \\
& = & 
\pi(\theta) 
\prod_{i=1}^n \ell_i(\theta_0)^{\sum_{j=1}^m w_{ij}}
\prod_{i=1}^n \left[
\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
\right]^{w_{ij}}
,
\end{eqnarray*}
therefore:
\begin{equation}
\label{eq:new_pool}
p_3(\theta|y) 
=
K \frac{\pi(\theta)}{\pi(\theta_0)}
\prod_{i=1}^n \left[
\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
\right]^{w_i(\theta)},
\end{equation}
where $K$ is a normalizing factor (dependent on~$y$) and the functions $w_i(\theta)$ are defined by $w_i(\theta_j)=w_{ij}$ for $j=1,\ldots,m$ and $w_i(\theta_0)=0$ conventionally.

As an opinion pooling rule, (\ref{eq:new_pool}) is more general than the log-linear pool~(\ref{eq:log_pool}) as it explicitly depends on~$\theta$ through the changing weights. Moreover, since the weights sum up to one for each~$\theta\not=\theta_0$, we have the equivalent expression:
$$
p_3(\theta|y) 
=
K
\prod_{i=1}^n \left[
\frac{\pi(\theta)\ell_i(\theta)}{\pi(\theta_0)\ell_i(\theta_0)}
\right]^{w_i(\theta)},
$$
showing that the pool does not change depending on whether the prior is incorporated before or after combining experts. In other words, (\ref{eq:new_pool}) is, again, externally Bayesian.

As in Section~\ref{sec:pool}, we recognize a Bayes rule-type expression: $p_3(\theta|y)\propto\pi(\theta){\cal L}_c(\theta,\mathbf{W})$, with:
\begin{equation}
\label{eq:super}
{\cal L}_c(\theta,\mathbf{W})
\equiv
\prod_{i=1}^n \left[
\frac{\ell_i(\theta)}{\ell_i(\theta_0)}
\right]^{w_i(\theta)}
,
\end{equation}
where~$\mathbf{W}$ denotes the $n\times m$ weight matrix with general element $w_{ij}$. We call~(\ref{eq:super}) a super composite likelihood (SCL). Note that the SCL evaluated at a particular hypothesis~$\theta_j$ boils down to a composite likelihood ratio,
\begin{equation}
\label{eq:super_ratio}
{\cal L}_c(\theta_j,\mathbf{W}) = 
\frac{L_c(\theta_j,\mathbf{w}_j)}{L_c(\theta_0,,\mathbf{w}_j)}
.
\end{equation}

To see that SCL~(\ref{eq:super}) is indeed a generalization of composite likelihood, assume that $\mathbf{w}_j=\mathbf{w}$ is the same for all hypotheses. This implies that (\ref{eq:super_ratio}) simplifies to ${\cal L}_c(\theta,\mathbf{W})=L_c(\theta,\mathbf{w})/L_c(\theta_0,\mathbf{w})$, the denominator of which is independent from~$\theta$ and can therefore be safely ignored for inference about~$\theta$. In this special case, SCL is equivalent to standard composite likelihood regardless of the chosen reference~$\theta_0$.

>>>>

\graphicspath{{.}{pics/}}

@@

\title{Composite Bayesian inference}
\date{}
\author{Alexis Roche\thanks{\url{alexis.roche@centraliens.net}}}

@@

\begin{abstract}
This note revisits the concept of composite likelihood from the perspective of probabilistic inference, and advocates a machine learning approach to tune the associated ``weights'', which stems naturally from a connection with the maximum entropy principle: the predictive distribution that maximizes entropy relative to a given prior and subject to multiple mean log-likelihood constraints is, up to a normalizing factor, the prior multiplied by a particular composite likelihood function, hence providing a ``composite'' extension of Bayes rule. We argue that composite Bayesian inference is a middle way between generative and discriminative approaches to statistical inference, and can be powerful in shallow learning problems, {\em e.g.}, in the context of transfer learning.
\end{abstract}

@@

Frequentist and Bayesian inference are based on the concept of likelihood, which relies on the existence of a statistical data-generating model that is both experimentally plausible and computationally tractable. Because this is challenging for complex data, other paradigms are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}, the minimum description length principle \cite{Grunwald-07} and, last but not least, discriminative models, which currently dominate the field of artificial intelligence and typically require supervised learning on large datasets -- these include a variety of deep learning \cite{Lecun-15,Goodfellow-16} and shallow learning techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00} and Gaussian processes \cite{Rasmussen-06}.

Composite likelihood (see \cite{Varin-11} and the references therein) is a semi-generative approach that extends the familiar notion of likelihood without requiring a full data-generating model. The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole. This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood. While maximum composite likelihood does not inherit the general property of maximum likelihood to yield asymptotically minimum-variance estimators, it may offer an excellent trade-off between computational and statistical efficiency.

In this note, composite likelihood is interpreted as a probabilistic opinion pool of ``agents'' making use of different pieces of information, or clues, extracted from the input data. Each agent acts as an isolated generative model-based statistician and expresses an opinion based on a specific clue in the form of a likelihood of the hidden variables. The opinion pool then compuptes a probability distribution on the hidden variables analogous to a Bayesian posterior, thereby extending Bayesian inference to problems where the complete likelihood is undefined.

We further show that a particular log-linear opinion pool yields the best possible predictive distribution in the sense of maximum entropy \cite{Grunwald-04}. This perspective entails a method to tune the weights of the different agents from training data as in a classical discriminative learning scenario.

@@

Let $Y$ an observable multivariate random variable with sampling distribution $p(y|x)$ conditional on some unobserved variable of interest, $X\in{\cal X}$, where ${\cal X}$ is a known set. Given an experimental outcome $y$, the likelihood is the sampling distribution evaluated at $y$, seen as a function of $x$:
$$
L(x) = p(y|x)
.
$$

This requires a plausible generative model which, for complex data, may be out of reach or involve too many nuisance parameters. A natural workaround known as {\em data reduction} is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the more convenient likelihood function:
$$
\ell(x) = p(z|x)
.
$$

Substituting $L(x)$ with $\ell(x)$ boils down to restricting the sample space, thereby  ``delegating'' statistical inference to an ``agent'' provided with partial information. While it is valid for such an agent observing~$z$ only to consider $\ell(x)$ as the likelihood function of the problem, the drawback is that $\ell(x)$ might yield too vague a prediction of~$X$ due to the information loss incurred by data reduction. To make the trick statistically more efficient, we may extract several features, $z_i=f_i(y)$ for $i=1,2,\ldots,n$, and try to combine the likelihood functions $\ell_i(x) = p(z_i|x)$ that they elicit.

If we regard the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. We will show in Section~\ref{sec:maxent} that one which is particularly well-suited to our case is the {\em log-linear opinion pool}: 
\begin{equation}
\label{eq:log_pool}
p_\lambda(x|y) = \frac{1}{Z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is some reference distribution or prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of real-valued weights so that the normalizing factor:
$$
Z_\lambda(y) = \int \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i} dx
$$
is finite\footnote{It is possible to have the weights depend on~$y$, however there is a strong rational for constant weights as we shall see in the sequel.}.

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding the form: $p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x)$, where the quantity:
\begin{equation}
\label{eq:comp_lik}
L^c_\lambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays the same role as a traditional likelihood function. We recognize in~(\ref{eq:comp_lik}) a general expression known as {\em marginal composite likelihood} \cite{Varin-11}. It shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, sometimes called {\em na\"ive Bayes} likelihood in the literature \cite{Ng-01}, which corresponds to the special case of all unitary weights, $\lambda_1=\ldots=\lambda_n= 1$. 
The general computational advantage of composite likelihood over the true likelihood is that it only requires to evaluate the marginal feature distributions rather than the joint distribution of all features.

\section{The problem of tuning composite likelihood weights}

There exist few general guidelines to choose the composite likelihood weights~$\lambda$ in~(\ref{eq:comp_lik}). Common practice is to impose nonnegative weights \cite{Varin-11} although this is not strictly required from the opinion aggregation viewpoint (the normalizing factor $Z_\lambda(y)$ in~(\ref{eq:log_pool}) only needs be finite for any~$y$ for the log-linear pool to yield a proper conditional distribution).

In cases where the features can be considered to be {\em exchangeable}, it is natural to pick uniform weights. The composite likelihood function is then a scaled version of the {\em na\"ive Bayes} likelihood, the common weight value being irrelevant to the location of the composite likelihood maximizer. It can be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the maximum composite likelihood estimator \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to reconcile the frequentist and Bayesian notions of uncertainty.

Another common recommandation for the composite likelihood weights is to sum up to one.

Going back to the log-linear pool underlying composite likelihood, Genest {\em et al} \cite{Genest-86b} showed that the only pooling operator that does not explicitly depend on~$x$ and preserves a property known as {\em external Bayesianity} is the log-linear opinion pool~(\ref{eq:log_pool}) with the additional constraint that the weights  sum up to one. External Bayesianity essentially means that it should not matter whether a prior on~$X$ is incorporated before or after combining opinions, provided that all agents agree on the same prior. Garg {\em et al} \cite{Garg-04} showed that the log-linear pool with unit weight sum also minimizes the average Kullback-Leibler (KL) divergence to the agent opinions.
The MaxEnt argument was already given by \cite{Wang-14}.

%Instead, redundancy between agents is assumed by default, and is effectively encoded by the unit sum constraint on weights\footnote{Nevertheless, features which are {\em known} to be mutually independent can be merged into a single feature. This results in increasing their weights in the log-linear pool.}.

The log-linear pool has, in particular, the 0/1 forcing property, meaning that if any agent assigns zero likelihood to a value~$x$, then $x$ has zero probability in the pool.

Unit sum weights = crap.

\section{Maximum entropy derivation of composite likelihood}
\label{sec:maxent}

The coordinator discards the priors used by the agents.

The MaxEnt argument was already given by \cite{Wang-14}.

% Accounting here for Xi'An comment on xianblog.wordpress.com
% The sum of the powers is constrained to be equal to one, even though
% I do not understand why the dimensions of the projections play no
% role in this constraint. Simplicity is advanced as an argument,
% which sounds rather weak…
This simple constraint is implied by the external Bayesianity requirement: as it turns out, the only aggregation operator which is both externally Bayesian and independent from $x$ boils down to plugging a composite likelihood with unit sum weights into Bayes rule, hence extending the classical notion of likelihood in Bayesian analysis.

@@

Composite likelihood may also be understood as a means to approximate the ``true'' likelihood function\footnote{By {\em true likelihood}, we mean the likelihood corresponding to a specified yet intractable generative model. In practice, such a model is obviously not required.} or, in the language of graphical models, the {\em message} that the data sends to the latent variable~$x$. Several ``clues'' are sent to~$x$ via different data features, and then integrated assuming {\em non-coalescent} emitting sources, meaning that statistical dependences between clues are treated as unknown, although not ignored. The whole idea is depicted by the factor graph\footnote{See \cite{Bishop-06} for a didactic introduction to factor graphs.} in Figure~\ref{fig:fgraph2}, where the latent variable is connected to multiple factors involving single clues (rather than a single factor involving multiple clues), thereby enabling efficient computations.

@@

\begin{figure}[!ht]
\begin{center}
\subfigure[Generative model]{\includegraphics[width=.35\textwidth]{fgraph1.pdf}\label{fig:fgraph1}}
\subfigure[Composite likelihood model]{\includegraphics[width=.6\textwidth]{fgraph2.pdf}\label{fig:fgraph2}}
\caption{Factor graph representations of a generative model (a) and its approximation using composite likelihood (b). In (a), the factor connecting the data~$y$ and the latent variable~$x$ is the true likelihood, $p(y|x)$. In (b), factors connecting~$y$ and the clues $z_1,z_2,\ldots$ represent feature extractions, $\alpha_i(y,z_i)=\delta[z_i-f_i(y)]$, while factors connecting the clues and~$x$ are scaled single-feature likelihoods, $\beta_i(z_i,x)=p(z_i|x)^{w_i}$. In both graphs, the factor shown in black is the prior $\pi(x)$.}
\label{fig:fgraph}
\end{center}
\end{figure}

@@

If one interprets the average divergence from agent opinions~(\ref{eq:avg_kl}) as a proxy for the divergence from ``the truth'', $p(x|y)$, composite likelihood can be seen as a variational approximation to the true likelihood. This corresponds to the intuitive notion that a consensus among sufficiently many experts should yield a reasonable guess. An essential difference with usual approximate inference methods \cite{Bishop-06,Minka-05} is that the likelihood function does not need to be computable owing to the use of a proxy. However, as we shall see in Section~\ref{sec:weights}, knowledge of the feature sampling distributions can be further exploited to optimize the likelihood approximation with respect to the composite likelihood weights.

\section{Conditional composite likelihood}
\label{sec:conditional}

The SCL derivation rests upon the definition of feature-based likelihood as $\ell_i(x)=p(z_i|x)$. As a straightforward  extension, $\ell_i(x)$ may be conditioned by an additional ``independent'' feature $z^c_i = f^c_i(y)$ considered as a predictor of the ``dependent'' feature, $z_i=f_i(y)$, yielding the more general form:
\begin{equation}
\label{eq:cond_feat_lik}
\ell_i(x) = p(z_i|x,z^c_i).
\end{equation}

Conditioning may be useful if it is believed that $z^c_i$ alone is little or not informative about $x$, but can provide relevant information when considered jointly with $x$, as in the case of regression covariates, for instance. Standard composite likelihood (\ref{eq:comp_lik}) then amounts to {\em conditional composite likelihood} \cite{Varin-11}, a more general form of composite likelihood also including Besag's historical {\em pseudo-likelihood} \cite{Besag-74}, which was a major breakthrough in computer vision.
